---
title: "Assignment 6: GLMs (Linear Regressions, ANOVA, & t-tests)"
author: "Abhay V Rao"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
getwd()
Chemistry_Data_Raw <- read.csv("../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", 
                               stringsAsFactors = FALSE)
#importing the raw dataset

library(lubridate) # loading lubridate to set date
library(tidyverse) #loading tidyverse to set theme, etc.
library(agricolae)# for HSD.test, etc.

Chemistry_Data_Raw$sampledate <- mdy(Chemistry_Data_Raw$sampledate)

class(Chemistry_Data_Raw$sampledate)

#2

mytheme <- theme_gray(base_size = 10) +
  theme(axis.text = element_text(color = "darkgrey"), 
        legend.position = "bottom")

theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The mean lake temperature does not change with depth across all lakes
Ha: The mean lake temperature changes with depth across all lakes


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
NTL_Wrangled_July <- Chemistry_Data_Raw %>%
  mutate(month= month(Chemistry_Data_Raw$sampledate))%>%
  filter(month=="7") %>%
  select(lakename, year4, daynum, depth, temperature_C)%>%
  filter(!is.na(temperature_C))

summary(NTL_Wrangled_July) # no NAs!

#5
NTL_regression_print <- 
  ggplot(NTL_Wrangled_July, aes(x = depth, y = temperature_C)) +
  geom_point() +
  geom_smooth(method = "lm", se=TRUE)+
  ylim(0,35)
print(NTL_regression_print) 

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: Temperature appears to reduce until a certain depth, where it starts
to level off, at about 8 meters. The correlation between depth and temperature
is apparent

7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
NTL_regression <- lm(data = NTL_Wrangled_July, temperature_C ~ depth)
summary(NTL_regression)

#The R-squared is 0.7387, and p-value is < 2.2e-16 
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: These results are based on 392 degrees of freedom. Depth explains about
73.8% of the variability in temperature. The result is statistically significant 
with a p-value of <2.2e-16, substantially lower than 0.05.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

LTER_AIC <- lm(data = NTL_Wrangled_July, temperature_C ~ year4 + daynum + 
              depth)

#Choose a model by AIC in a Stepwise Algorithm
step(LTER_AIC)
LTER_AIC_model <- lm(data = NTL_Wrangled_July, temperature_C ~ year4 + daynum + 
              depth) 
summary(LTER_AIC_model)

#the AIC indicates that all the explanatory variables are significant 

#10
NTL_regression2 <- lm(data = NTL_Wrangled_July, temperature_C ~ year4+ daynum + depth)
summary(NTL_regression2)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC suggests that the variables to consider are year4, depth and daynum.
These variables explain 74.12% of the variability in temperatures. The p-value
remains the same, so the results are significant. Considering we have additional
insight on variability without an increase in p-value, this is an improved
model.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

# null hypothesis: there is no difference in means

Lake_Temp_anova <- aov(data = NTL_Wrangled_July,  temperature_C ~ lakename)
summary(Lake_Temp_anova)

# P-value of <2e-16 indicates null hypothesis is invalid.

# Next, formatting ANOVA as lm
Lake_Temp_anova2 <- lm(data = NTL_Wrangled_July,  temperature_C ~ lakename)
summary(Lake_Temp_anova2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Based on this analysis, there is a significant difference in means
between lakes, evidenced from the p-value of on <2e-16.The lm-based analysis also
indicates a substantial difference in means as compared to the intercept term.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

NTLplot <- ggplot(NTL_Wrangled_July, aes(x = depth, y =temperature_C, color=lakename)) +
  geom_point(alpha=0.5) +
  geom_smooth(method = "lm",color="BLACK", se= FALSE)+
  ylim(0,35)+
  facet_wrap(vars(lakename), nrow = 2)
print(NTLplot)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(Lake_Temp_anova)#assessing difference between means 

TreatmentGroups_LakeTemp <- HSD.test(Lake_Temp_anova,"lakename", group = TRUE)
TreatmentGroups_LakeTemp #grouping lakes which have similar means

#visualizing these groupings
HSDGroups.plot <- ggplot(NTL_Wrangled_July, aes(x = lakename, y = temperature_C)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  stat_summary(geom = "text", fun = max, vjust = -1, size = 3.5,
               label = c("a", "ab", "e", "de", "c", "c", 
                         "de", "bc", "d")) +
  labs(x = "Lake Name", y = "Temperature (C)") +
  ylim(0, 35)
print(HSDGroups.plot)
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Peter, Paul and Ward Lake all fall into group "c" and can be considered
to have the same mean statistically. No lake has a mean that is distinct from
ALL other lakes.


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We can do a two sample t-test to see the difference between the means of the two
lakes - ANOVAs are required in those cases where we have more than two samples.


